Predictive State Geometry via Cantor Embeddings and Wasserstein Distance

### 2022-02-06 paper/manuscript.tex
JPC: Done with edit pass. Ready to submit to KDD.

%%
%% The abstract is a short summary of the work to be presented in the % article.
\begin{abstract}
  Predictive states for stochastic processes are a non-parametric and
  interpretable construct with relevance across a multitude of modeling
  paradigms. Recent progress on the self-supervised reconstruction of predictive
  states from time-series data has focused on the use of reproducing kernel
  Hilbert spaces. Here, we examine how Wasserstein distances may be used to
  detect predictive equivalences in symbolic data. We construct a
  finite-dimensional embedding of sequences based on the Cantor set, and use
  distances in this embedding to compute Wasserstein distances between
  distributions over sequences (``predictions''). We show that analyzing the
  resulting geometry (via hierarchical clustering and dimension reduction)
  provides insight into the temporal structure of processes ranging from the
  relatively simple (\emph{e.g.}\ hidden Markov models) to the very complex
  (\emph{e.g.}\ indexed grammars).
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe % the work
%being presented. Separate the keywords with commas.
\keywords{time series, predictive states, wasserstein distance, hierarchical clustering}

@@

\section{Introduction}
Suppose that we have a finite sequence $x_1 \dots x_L$ of categorical
observations drawn from a temporal process. We may suppose that the process is
stationary (time-translation-invariant) and ergodic (has a tendency to explore
all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
observed information the behavior of the next $n$ observations. If we suspect
that temporal correlations of the process do not matter much beyond $k$ symbols,
then we can use the data we have to take all subsequences of length $n+k$,
splitting each subsequence into words of length $k$ and $n$ respectively. From
these ``past/future'' pairs we could construct an empirical conditional
distribution
\begin{equation}\label{eq:conditional}
  \hat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
  \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
\end{equation}
where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
\dots x_L$.

But suppose now that we do not know how long-range the temporal dependencies in
our process can stretch. Even very simple stochastic processes can have infinite
Markov order, indicating potential long-term dependence of future observations
on the past \cite{Uppe97a}. Given sufficient data, it would be desirable to take
pasts of arbitrary length, and converge towards some prediction conditioned on
the \emph{infinite} past:
\begin{equation}\label{eq:convergence}
  {P}_{x_{1}\dots x_{n}|\overleftarrow{x}}
  = \lim_{k\rightarrow\infty}
  \hat{P}_{x_{1}\dots x_{n}|x_{-k}\dots x_0}
\end{equation}
where $\overleftarrow{x}$ denotes an infinite sequence $\overleftarrow{x} =
(\dots,x_{-1},x_0)$ of observations stretching into the past. This mathematical
ideal is known to previous literature as the \emph{causal} or \emph{predictive
state} \cite{Crut88a,Jaeg00a}. Formally, the conditional predictions
${P}_{x_{1}\dots x_{n}|\overleftarrow{x}}$ for all forecast lengths $n$ together
describe a \emph{probability measure} over future sequences $\overrightarrow{x}
= (x_1,x_2,\dots)$; the predictive state is this measure. We will denote it
simply by $P_{\overleftarrow{x}}$.

Predictive states have been utilized for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a} and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}, and a deep mathematical
theory of predictive state inference has been correspondingly developed
\cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
this, universal conditions for predictive state convergence were not given until
recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
stationary process, and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} will converge for any $n$ and a probability-1 subset of
pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$.

The general goals of predictive state analysis are typically threefold
\cite{Shal01a}. The first is to understand the overall structure of how the
predictive states relate to one another geometrically, and possibly use this
geometry to classify pasts based on equivalence of their predictive states. The
second is to actually reproduce the prediction to a specified accuracy. The
third is to understand the dynamics of how predictions evolve under a stream of
new observations. We will focus in this paper on the first, as it is a crucial
building block to achieving the other two.

Recent attempts at reconstructing the geometry of predictive states have focused
on embedding them in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}, to great effect largely because
convergence in Hilbert spaces generated by universal kernels is equivalent to
convergence in distribution \cite{Srip10a}. That is, these embeddings allow
accurate representations of predictive states because they respect the product
topology of sequences in the same way that predictive states themselves do.

Understanding the source of the power of RKHS methods in predictive state
analysis frees us to consider other options. In this paper, we will embed
symbolic sequences in a one-dimensional space that has the same topology as the
product space. This embedding is inspired by the fractal Cantor set
\cite{Kurk03a}. Predictive states can then be thought of as distributions in
this one-dimensional space. We then use the Wasserstein distance to compute the
geometry between predictive states, which can be computed with a closed-form
integral for one-dimensional distributions. This works as an alternative to
RKHS-based distances because the Wasserstein distance also reproduces the
topology of convergence in distribution \cite{Pane19a}. The resulting distance
matrix can be used to find low-dimensional embeddings \cite{Borg05a} of the
geometry or hierarchical clusterings \cite{Mull11a} of the predictive states.
The latter, in particular, when combined with our fractal embedding, provides a
highly interpretable visualization of the predictive state space.

\section{Example processes}
The methods in this paper are intended to be applied toward stationary and
ergodic stochastic processes which produce categorical time-series data. For our
intents we can think of a stochastic process as a collection of probability
distributions $\Prob{\mu}{x_1\dots x_L}$ over any finite, contiguous sequence,
taking values in a finite set $\mathcal{X}$. Formally this describes a measure
$\mu$ over the set of all bi-infinite sequences $(\dots,x_{-1},x_0,x_1,\dots)\in
\mathcal{X}^{\mathbb{Z}}$. These sorts of processes can be generated by a number
of systems with widely varying complexity. Most popularly studied are those
methods often characterized as having some degree of ``finite memory'': Markov
models, hidden Markov models, and observable operator models (also termed
generalized hidden Markov models) \cite{Uppe97a,Jaeg00a}. Beyond these one can
also generate processes using probabilistic grammars, such as probabilistic
context-free and indexed grammars \cite{Gema00a}. Additionally, coarse-grained
data from dynamical systems (such as the logistic map) can display behavior
varying widely in complexity \cite{Crut92c}. In this paper we will refer back
frequently to the following example processes:
\begin{enumerate}
  \item The \emph{even process} can be generated by repeatedly tossing a coin
  and writing down a $0$ for every tail and $11$ for every head. The process is
  essentially random except for the constraint that $1$'s can only appear in
  contiguous blocks of even size. The even process has infinite Markov order but
  can be generated by a two-state hidden Markov process \cite{Crut01a}.
  \item The $a^n b^n$ \emph{process} can be generated by choosing a random
  integer $n\geq 1$ (we will suppose via a Poisson process) and writing $n$
  $\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
  repeating this process indefinitely. The result will be sequences where any
  contiguous block of $\mathtt{a}$'s is followed by a block of $\mathtt{b}$'s of
  equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be modeled by any
  hidden Markov model, though it is a simple example of a probabilistic
  context-free grammar \cite{Hopc06a}.
  \item The $x+f(x)$ \emph{process} is a probabilistic context-free grammar
  modeling simple mathematical expressions. It has terminal symbols
  $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\ \mathtt{+}\ ,\ \mathtt{f}\ ,\
  \mathtt{x}\}$ and non-terminals $\{A,B,C\}$, and starts with a sequence osf
  $A$'s with the production rules:
  \begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  \end{align*}
  \item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed grammar
  \cite{Hopc06a} which is analagous to $\mathtt{a}^n \mathtt{b}^n$ except after
  writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block
  of $\mathtt{c}$'s of length $n$.
  \item The \emph{Feigenbaum process} is generated by sampling from the time
  series of the logistic map at critical parameter $r\approx 3.56995$:
  \begin{equation*}
    y_{t+1} = r y_t(1-y_t)
  \end{equation*}
  and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
  \frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
  Alternatively, we can generate this process by starting with a single $0$ and
  executing the replacements $0\mapsto 11$ and $1 \mapsto 01$ consecutively. The
  result of the Feigenbaum process is indexed-context free \cite{Crut92c}.
\end{enumerate}

\section{Cantor-embedding sequences}
\begin{figure*}[h]
  \centering
  \includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
  \caption{ Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
    $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
    Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
    past and future respectively. The symbol on the $x$ ($y$) axis indicate that
    all points above (to the right of) that symbol have a past (future) whose
    most recent observation is that symbol. Though not marked, further
    proportional subdivisions of each segment of the axes would indicate the
    value of the second, third, \emph{etc.} symbols. For instance, one can read
    from the $\mathtt{x+f(x)}$ fractal that any past which ends in $\mathtt{f}$
    must be paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$. }
    \label{fig:cantor}
  \Description{Four fractals are displayed, respectively labeled as the even,
  $\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
  $\mathtt{x+f(x)}$ processes.}
\end{figure*}

@@

We can exploit this self-similarity in an interesting way by constructing a
mapping between sequence space and the celebrated Cantor set (or one of its
generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
can associate some unique integer between $0$ and $|\mathcal{X}|-1$ inclusive;
call this $J(x)$. Then there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
[0,1]$ which maps every sequence to a positive real number:
\begin{equation}
  C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
\end{equation}
For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
sequence to a point the traditional Cantor set fractal. For a finite sequence of
length $L$, truncate the sum at $k=L$.

Remarkably, the embedding $C$ has the property that for any continuous function
$f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
is continuous on $\mathcal{X}^\mathbb{N}$; further, if $F$ is continuous on
$\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
Thus the embedding $C$ respects the basic structure of the product topology
\cite{Kurk03a}.

Stationary processes, due to their time-translation invariance, inherit the
fractal temporality of sequence space. This can be neatly visualized: given a
length-$L$ sample $x_1\dots x_L$, and some $n,k>0$, take a sliding window of
pasts and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.
For each past-future pair, compute the truncated Cantor embeddings on the
\emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
be plotted as $(x,y)$-values on a scatter plot. The fractal which emerges
contains, in essence, all information necessary to understand the temporal
structures of the process. See Fig. \ref{fig:cantor} for examples and guidance
on how to interpret the visualization.

It is worth noting that for $|\mathcal{X}|>2$, the embedding $C$ also introduces
additional structure which may or may not be desirable. The association of each
symbol $x$ with an integer $j_x$ endows an ordinal structure on the set
$\mathcal{X}$; this ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later we will examine higher-dimension
embeddings which do not have this ordinal aspect; however, they will come at the
cost of increased computational complexity for the Wasserstein distance
computation, so for now we will assume that ordinal artifacts are either desired
or sufficiently tolerable as to not outweight the computational benefits of
working in one dimension.

@@

The Cantor fractals in Fig. \ref{fig:cantor} can be thought of as probability
distributions. We can consequently interpret a vertical slices of the fractal,
located at horizontal position $C(\overleftarrow{x})$, as visualizing the
predictive state $P_{\overleftarrow{x}}$ as a distribution over Cantor-embedded
futures, $C(\overrightarrow{x})$.

For example, by analyzing the Cantor fractal of the even process, one may notice
that there are effectively only 2 distinct predictive states---every vertical
column is just one of two types. This corresponds with the 2 states of the
hidden Markov model which generates the even process.

This visualization allows us to see how predictive states distribute their
probability over the intrinsic geometry of potential futures. We can compare
predictive states not only on how much their supports overlap, but on how
geometrically close their supports are to one another. For the $\mathtt{a}^n
\mathtt{b}^n$ process, for example, we can see that the first few columns
(corresponding to pasts of the form $\dots \mathtt{b a}^n$ for some $n$) are
inherently similar to one another, though they are shifted upwards the closer to
the axis they are (which corresponds to the increasing number of $\mathtt{b}$'s
in the predicted future as $n$ increases).

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$, $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by
\begin{equation*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
\end{equation*}
where $\Gamma(\mu,\nu)$ is the set of all measures on
$\mathcal{M}\times\mathcal{M}$ whose left and right marginals are $\mu$ and
$\nu$ respectively. This can be interpreted as the minimal cost for ``shifting''
the probability mass from one distribution to match shape of the other.

$W(\mu,\nu)$ the solution to a constrained linear optimization. As a function of
distributions, $W(\mu,\nu)$ is continuous with respect to convergence in
distribution; in fact, convergence under the Wasserstein distance is equivalent
to convergence in distribution on compact spaces. This makes $W(\mu,\nu)$ ideal
for measuring geometry between predictive states, since empirical estimates of
these are known to converge in distribution.

When $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
respectively be the cumulative distributions functions of $\mu$ and $\nu$. Then
\begin{equation}
  W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
\end{equation}
This closed-form solution is considerably faster to compute than the linear
optimization required for arbitrary metric spaces. Because the Cantor embedding
embeds the space of sequences directly into $[0,1]$, we can utilize this
formula.

The combination of the Cantor embedding and the Wasserstein distance lays out a
fairly straightforward programme for analyizing a categorical time series:
\begin{enumerate}
  \item Apply Algorithm \ref{alg:cantorwass} to the data stream $x_1\dots x_L$
  for a specified past length $k$ and future length $n$, retrieving the set of
  unique observed pasts, the empirical Cantor distributions corresponding to
  each past, and the matrix of Wasserstein distances computed from these
  distributions.
  \item Use the Wasserstein distance matrix to perform additional methods of
  geometric data analysis, such as hierarchical clustering \cite{Mull11a} and
  multidimensional scaling \cite{Borg05a}, in order to elucidate the relative
  geometry of the predictive states.
\end{enumerate}
In the next two sections we examine the results of this approach.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the % article.
\begin{abstract}
  Predictive states for stochastic processes are a nonparametric and
  interpretable construct with relevance across a multitude of modeling
  paradigms. Recent progress on the self-supervised reconstruction of predictive
  states from time-series data focused on the use of reproducing kernel
  Hilbert spaces. Here, we examine how Wasserstein distances may be used to
  detect predictive equivalences in symbolic data. We construct a
  finite-dimensional embedding of sequences based on the Cantor set and use
  distances in this embedding to compute Wasserstein distances between
  distributions over sequences (``predictions''). We show that analyzing the
  resulting geometry via hierarchical clustering and dimension reduction
  provides insight into the temporal structure of processes ranging from the
  relatively simple (\emph{e.g.}, finite-state hidden Markov models) to the
  very complex (\emph{e.g.}, infinite-state indexed grammars).
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe % the work
%being presented. Separate the keywords with commas.
\keywords{time series, predictive states, Cantor set, Wasserstein distance, hierarchical clustering}

@@

\section{Introduction}
Suppose that we have a finite sequence $x_1 \dots x_L$ of categorical
observations drawn from a temporal process. We may suppose that the process is
stationary (time-translation invariant) and ergodic (explores
all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
observed information the behavior of the next $n$ observations. If we suspect
that the process' temporal correlations do not matter much beyond $k$ symbols,
then we take our data to be all subsequences of length $n+k$,
splitting each subsequence into words of length $k$ and $n$ respectively. From
these ``past/future'' pairs we can construct an empirical conditional
distribution
\begin{align*}
  \widehat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
  \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
  ~,
\end{align*}
where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
\dots x_L$.

Suppose, though, that we do not know how long-range the process' temporal
dependencies can stretch. Even very simple stochastic processes can have
infinite Markov order, indicating potential long-term dependence of future
observations on the past \cite{Uppe97a}. Given sufficient data, it would be
desirable to take pasts of arbitrary length and converge towards a prediction
conditioned on the \emph{infinite} past:
\begin{align}
  {P}_{x_{1}\dots x_{n}|\overleftarrow{x}}
  = \lim_{k\rightarrow\infty}
  \widehat{P}_{x_{1}\dots x_{n}|x_{-k}\dots x_0}
\label{eq:convergence}
\end{align}
with the infinite sequence $\overleftarrow{x} = (\dots,x_{-1},x_0)$ of
observations stretching into the past. This mathematical ideal is known as the
\emph{causal} or \emph{predictive state} \cite{Crut88a,Jaeg00a}. Formally, the
conditional predictions ${P}_{x_{1}\dots x_{n}|\overleftarrow{x}}$ for all
forecast lengths $n$ together describe a \emph{probability measure} over future
sequences $\overrightarrow{x} = (x_1,x_2,\dots)$; the predictive state is this
measure. We denote it simply $P_{\overleftarrow{x}}$.

Predictive states are employed for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
theory of predictive-state inference has been correspondingly developed
\cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
this, universal conditions for predictive-state convergence were not given until
recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
stationary process and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$.

The broad goals of predictive-state analysis are threefold \cite{Shal98a}. The
first is to understand the overall structure of how the predictive states
relate to one another geometrically and, possibly, use this geometry to
classify pasts based on equivalence of their predictive states. The second is
to actually reproduce the prediction to a specified accuracy. The third is to
understand the dynamics of how predictions evolve under a stream of new
observations. The following focuses on the first, as it is a crucial building
block to achieving the other two.

Recent attempts to reconstruct the geometry of predictive states embedded them
in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do.

Understanding the source of the power of RKHS methods in predictive-state
analysis frees us to consider other options. The following embeds
symbolic sequences in a one-dimensional space that has the same topology as the
product space. This embedding is inspired by the fractal Cantor set
\cite{Kurk03a}. Predictive states can then be thought of as distributions in
this one-dimensional space. We then use the Wasserstein distance to compute the
geometry between predictive states, which is determined by a closed-form
integral for one-dimensional distributions. This operates as an alternative to
RKHS-based distances since the Wasserstein distance also reproduces the
topology of convergence in distribution \cite{Pane19a}. The resulting distance
matrix then is used to find low-dimensional embeddings \cite{Borg05a} of the
geometry or hierarchical clusterings \cite{Mull11a} of the predictive states.
When combined with the fractal embedding, the latter, in particular, provides a
highly interpretable visualization of the predictive-state space.

\section{Example processes}

The methods here are intended to be applied to stationary and ergodic
stochastic processes that generate categorical time-series data. For these
purposes we consider a stochastic process to be a collection of probability
distributions $\Prob{\mu}{x_1\dots x_L}$ over any finite, contiguous sequence,
taking values in a finite set $\mathcal{X}$. Formally, this describes a measure
$\mu$ over the set of all bi-infinite sequences $(\dots,x_{-1},x_0,x_1,\dots)
\in \mathcal{X}^{\mathbb{Z}}$.

These processes are generated by a number of systems with widely varying
complexity. Most popularly studied are those often characterized as having a
degree of ``finite memory'': \emph{Markov chains}, \emph{hidden Markov chains},
and \emph{observable operator models} (also termed \emph{generalized hidden
Markov chains}) \cite{Uppe97a,Jaeg00a}. Beyond these, one can also generate
processes using probabilistic grammars, such as probabilistic context-free and
indexed grammars \cite{Gema00a}. Additionally, coarse-grained data from chaotic
dynamical systems---such as the \emph{logistic map}---display behavior varying
widely in complexity \cite{Crut92c}.

We refer back frequently to the following example processes:
\begin{enumerate}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}.
\item The $a^n b^n$ \emph{process} can be generated by choosing a random
	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
	repeating this procedure indefinitely. This results in sequences where any
	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
	by any finite hidden Markov chain, though it is a simple example of a
	probabilistic context-free language \cite{Hopc06a}.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
	write a block of $\mathtt{c}$'s of length $n$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}.
\end{enumerate}

\begin{figure*}[h]
\centering
\includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
	that all points above (to the right of) that symbol have a past (future)
	whose most recent observation is that symbol. Though not marked, further
	proportional subdivisions of each segment of the axes indicate the value of
	the second, third, \emph{etc.} symbols. For instance, one can read from the
	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
	}
\label{fig:cantor}
\Description{Four fractals are displayed, respectively labeled as the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes.
	}
\end{figure*}

\section{Cantor-embedding sequences}

@@

We exploit this self-similarity in an interesting way by constructing a mapping
between sequence space and the celebrated \emph{Cantor set} (or one of its
generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
associate a unique integer between $0$ and $|\mathcal{X}|-1$ inclusive; call
this $J(x)$. Then, there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
[0,1]$ that maps every sequence to a positive real number:
\begin{align*}
  C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
  ~.
\end{align*}
For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
sequence to a point the traditional Cantor set fractal. For a finite sequence of
length $L$, truncate the sum at $k=L$.

Remarkably, the embedding $C$ has the property that for any continuous function
$f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
is continuous on $\mathcal{X}^\mathbb{N}$. Further, if $F$ is continuous on
$\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
Thus, the embedding $C$ respects the basic structure of the product topology
\cite{Kurk03a}.

Stationary processes, due to their time-translation invariance, inherit the
fractal temporality of sequence space. This can be easily visualized: Given a
length-$L$ sample $x_1\dots x_L$, and $n,k>0$, take a sliding window of pasts
and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.  For
each past-future pair, compute the truncated Cantor embeddings on the
\emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
be plotted as $(x,y)$-values on a scatter plot. The fractal that emerges
contains, in essence, all information necessary to understand a process'
temporal structures. See Fig. \ref{fig:cantor} for examples and guidance on how
to interpret the visualization.

Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
structure that may or may not be desired. Associating each symbol $x$ with an
integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
embeddings that do not have this. However, they come at the cost of increased
computational complexity when determining the Wasserstein distance. So, for now,
we assume that ordinal artifacts are either desired or sufficiently tolerable
as to not outweigh the computational benefits of working in one dimension.

@@

Figure \ref{fig:cantor}'s Cantor fractals represent probability distributions:
We interpret a vertical slice of the fractal, located at horizontal position
$C(\overleftarrow{x})$, as visualizing the predictive state
$P_{\overleftarrow{x}}$ as a distribution over Cantor-embedded futures
$C(\overrightarrow{x})$.

For example, by examining the even process' Cantor fractal, one notices that
there are effectively only $2$ distinct predictive states---every vertical
column is just one of two types. This corresponds with the $2$ states of the
hidden Markov model that generates the even process.

This allows us to see how predictive states distribute their probability over
the intrinsic geometry of potential futures. We compare predictive states not
only on how much their supports overlap, but on how geometrically close their
supports are to one another. For the $\mathtt{a}^n \mathtt{b}^n$ process, for
example, we see that the first few columns (corresponding to pasts of the form
$\dots \mathtt{b a}^n$ for some $n$) are inherently similar to one another,
though they are shifted upwards the closer to the axis they are. (The latter
corresponds to the increasing number of $\mathtt{b}$s in the predicted future
as $n$ increases.)

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by:
\begin{align*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
   ~,
\end{align*}
where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
It is the minimal cost to ``shift'' the probability mass from one distribution
to match shape of the other.

$W(\mu,\nu)$ is the solution to a constrained linear optimization. As a
function of distributions, $W(\mu,\nu)$ is continuous with respect to
convergence in distribution. In fact, convergence under the Wasserstein
distance is equivalent to convergence in distribution on compact spaces. This
makes $W(\mu,\nu)$ ideal for measuring geometry between predictive states,
since empirical estimates of these are known to converge in distribution.

When $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
respectively be the cumulative distributions functions of $\mu$ and $\nu$.
Then:
\begin{align*}
  W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
  ~.
\end{align*}
This closed-form solution is considerably faster to compute than the linear
optimization required for arbitrary metric spaces. Since the Cantor embedding
embeds the space of sequences directly into $[0,1]$, we can directly employ
this formula.

Combining the Cantor embedding and Wasserstein distance leads to a
straightforward program for analyzing categorical time series:
\begin{enumerate}
\item Apply Algorithm \ref{alg:cantorwass} to the data stream $x_1\dots x_L$
	for a specified past length $k$ and future length $n$, retrieving the (i)
	set of unique observed pasts, (ii) empirical Cantor distributions
	corresponding to each past, and (iii) matrix of Wasserstein distances
	computed from these distributions.
\item To elucidate the relative geometry of the predictive states, use the
	Wasserstein distance matrix to perform additional methods of geometric data
	analysis, such as hierarchical clustering \cite{Mull11a} and
	multidimensional scaling \cite{Borg05a}.
\end{enumerate}
The next two sections examine the results of this approach.
